---
title: "Weekly 2: Coffee Ratings"
author: "[Name]"
format:
  html:
    embed-resources: true
editor: visual
---

## Overview

You want to know which producers grow the best-rated coffee.

That sounds like a `group_by()` and a `summarise()`. The problem is that `producer` was typed by different people in different ways, so the same producer can appear under several labels, and some labels are not producers at all.

Your job is to work out how much of that you can fix, how much you shouldn't, and what that does to the answer.

A few things:

1.  Add `#| echo: false` to your code blocks.
2.  Use a new block when you move to a new task.
3.  Code goes in the blocks. **Answers go outside them**, as prose. A comment inside a code chunk is not an answer.
4.  Keep the original `producer` column. Every cleaned version goes in a **new** column.

Due Monday at 8pm.

## Your data

Import `coffee_ratings.csv` and inspect it.

```{r}
#| warning: false
library(tidyverse)

ratings <- read_csv("../Data/Raw/coffee_ratings.csv")

glimpse(ratings)
```

## Before you start

> What does one row represent?
>
> A coffee sample

> Is there a column that uniquely identifies a row? How would you check?
>
> Nope, you look to see if there is a col with only unique values

> Is `producer` a key? What does it mean that the same producer can appear on many rows?
>
> Producer is not a key because several rows can share one, if it appears multiple times that means its multiple samples from the same producer

## How many producers?

Count the distinct values of `producer`.

```{r}
#| warning: false
ratings %>%
  summarise(
    num_of_producers = n_distinct(producer)
  )
```

> How many distinct values are there?
>
> 692

> How many rows have no producer recorded? Does your distinct count include "missing" as one of its values? How would you check?
>
> ```         
> Yes
> ```
>
> This is included in distinct
>
> ```{r}
> #| warning: false
> ratings %>%
>   summarise(
>     num_of_producers = sum(is.na(producer))
>   )
> ```

> Why shouldn't you read that number as the number of distinct producers?
>
> Because there are a lot of secret null values

## Normalize the representation

Create a new column, `producer_clean`, that makes the producer labels consistent in case and whitespace. Keep `producer`.

```{r}
#| warning: false
ratings <- ratings %>%
  mutate(
    producer_clean = toupper(str_squish(producer))
  )

```

> How many distinct values does `producer_clean` have?
>
> 666 - The Anti Christ
>
> ```{r}
> #| warning: false
> ratings %>%
>   summarise(
>     num_of_producers = n_distinct(producer_clean)
>   )
> ```

> Is that the number of producers now? What does the reduction establish, and what does it *not* establish?
>
> No, It cuts a few duplicates, but it still has thigns like null

## Which labels actually merged?

A single number doesn't tell you what happened. Build a table showing every `producer_clean` value that more than one raw `producer` value was combined into, along with the raw values that went into it.

```{r}
#| warning: false
ratings %>%
  group_by(producer_clean) %>%
  summarise(
    num_of_raw_values = n_distinct(producer, na.rm = TRUE)
  ) %>%
  filter(num_of_raw_values > 1)
```

> How many cleaned labels absorbed more than one raw label?
>
> 22

> Look through them. Is every merge one you would stand behind?
>
> Yea it look like it

> Is this table more useful than knowing the number of labels went down? Why?
>
> It is because it shows how it's working giving us a deeper understanding of what we're manipulating

## Does it change anything?

Compare the La Plata coffees before and after normalizing. For each version, report the number of coffees and the mean and median `total_cup_points`.

```{r}
#| warning: false
ratings %>%
  filter(producer == "La Plata") %>%
  summarise(
    mean_total = mean(total_cup_points),
    median_total = median(total_cup_points),
    total = sum(total_cup_points)
  )

ratings %>%
  filter(producer_clean == "LA PLATA") %>%
  summarise(
    mean_total = mean(total_cup_points),
    median_total = median(total_cup_points),
    total = sum(total_cup_points)
  )
```

> What changed more: the number of coffees or the average score?
>
> Number of coffees, it changed dramatically more than the averages

> Suppose you had reported how many coffees La Plata produced using the raw column. How far off would you have been?
>
> By about 700

## What case and whitespace didn't catch

Some labels still differ only in punctuation, i.e., a period, a comma, a slash.

Sort the distinct values of `producer_clean` and read through them. Near-duplicates end up next to each other.

```{r}
#| warning: false
ratings <- ratings %>%
  mutate(
    producer_clean = str_remove_all(producer_clean,
                                    "[[:punct:]]") 
  )
```

> Find at least **three** pairs of labels that differ only in punctuation.
>
> ```         
> AGRICOLA EL PILAR, S.A. | AGRICOLA EL PILAR, SA.
> AGRICOLA NUEVA GRANADA, S.A | AGRICOLA NUEVA GRANADA, S.A.   
> ConquistaMorito | Conquista/Morito 
> ```
>
> ```{r}
> #| warning: false
> ratings %>%
>   filter(!is.na(producer)) %>%
>   mutate(producer_clean = str_remove_all(producer, "[[:punct:]]")) %>%
>   group_by(producer_clean) %>%
>   summarise(
>     producers = paste(unique(producer), collapse = " | ")
>   ) %>%
>   filter(str_detect(producers, fixed(" | ")))
> ```

Now use exact replacement to make each pair consistent. Change only the values you identified and nothing else.

```{r}
#| warning: false

ratings %>%
  mutate(
    producer = str_replace_all(
      producer,
      c(
        "AGRICOLA EL PILAR, S.A." = "AGRICOLA EL PILAR, SA.",
        "AGRICOLA NUEVA GRANADA, S.A" = "AGRICOLA NUEVA GRANADA, S.A.",
        "ConquistaMorito" = "Conquista/Morito"
      )
    )
  )
```

> For each change, say why you're confident it's the same producer.
>
> 1: "AGRICOLA EL PILAR, S.A." = "AGRICOLA EL PILAR, SA."\
> This is literally the same spelling but the acronym is missing a period
>
> 2: "AGRICOLA NUEVA GRANADA, S.A" = "AGRICOLA NUEVA GRANADA, S.A."
>
> Same situation, an acronym is missing a period
>
> 3: "ConquistaMorito" = "Conquista/Morito"
>
> I think these are the same but only separated but a slash

> How long did it take to find those by reading? How many labels did you have to look at?
>
> I just used r lol

## The top producers

Using `producer_clean`, find the ten producers with the most coffees in the data.

```{r}
#| warning: false
ratings%>%
  group_by(producer_clean) |>
  summarise(
    top_producers = n()
  ) |>
  slice_max(order_by = top_producers, n = 10)
```

Look at that list carefully.

> Is every entry in it a producer? If not, which ones aren't, and what do they mean?
>
> no, some are NA or "Several"

> Is any single producer split across more than one row of the list?
>
> Yes

Now look at these four raw labels:

- Ipanema Agricola
- Ipanema Agricola S.A
- Ipanema Agrícola SA
- Ipanema Coffees

> Should these be combined into one producer? **Don't recode them.** Explain whether combining them would be normalization or classification, and what evidence you'd need before doing it.
>
> They should be combined, it is clear they are all variations of the same name

> If you did combine them, what would happen to your top-ten list?
>
> Ipanema would odviously blow up and have a high ranking

## What this means for your question

> A classmate says: "I reduced the producer column from 692 labels to 666, so the cleaned column is better." Is that justified? Explain.
>
> Yea because then it has less noise from spelling variations

> You set out to find which producers grow the best-rated coffee. Given everything above, what would you need to tell someone before they trusted your answer?
>
> I'd tell them that the data has been cleaned first

## Disclosure appendix

> What did you try first? Include the code you wrote and the actual error message, or a clear description of the problem if there was no error.

> If you used an AI system: the exact prompt, exactly what it returned, what it got wrong or left unclear, and what you changed.

> If you didn't use one, say so.
